pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-13 article.tex
To and including '2 Description'

% Affiliations
\usepackage{authblk}
\title{pirouette: Determining the error BEAST2 makes in inferring a phylogeny}
\author[1]{Rich\`el J.C. Bilderbeek}
\author[1]{Giovanni Laudanno}
\author[1]{Rampal S. Etienne}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}

@@

Although many possible tests can be performed, 
we present here only five example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will estimate the error made by BEAST2 when 
using a birth-death tree prior, in recovering an original tree 
generated under a Yule generative prior.

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & \textbf{Possible values} \\ 
    \hline
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model & Yule, BD, CBS, CCP, CEP \\
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates & strict, relaxed log-normal \\
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model & JC, HKY, TN93, GTR \\
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs & $m_{r} \in \mathbb{R}_{>0}$\\
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree & any combination of a, c, g, t \\
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common ancestor & see documentation \\
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup & see documentation \\
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model & Generative, Candidate \\
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is used & Always, Best candidate\\
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error & nLTT, $|\gamma|$ \\
    $\mathit{t_{m}}$ & twin tree prior & Sets the tree prior for the twin tree & Yule, BD \\
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. BD = birth death, CBS = coalescent
    Bayesian Skyline, CCP = coalescent constant-population, CEP = coalescent
    exponential-population, JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), HKY = Hasegawa, Kishino and 
    Yano (\cite{hasegawa1985dating}), TN93 = \cite{tamura1993estimation}, GTR = Generalised time-reversible 
    model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a) simulated under the selected generative prior $\mathit{P}$, that is converted to an alignment (2a) according to the alignment model $\mathit{A}$. The user supplies one or more experiments. Every experiment $\mathit{X}$ is defined as a combination of an inference model $\mathit{I}$ and the conditions $\mathit{C}$ under which it is actually used in the inference. Based on the experiments and possibly the alignment (2a), experiments are selected to be actually used in the next step (3a).
    Selected experiments use the alignment (2a) and their inference model $\mathit{I}$ to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). The posterior trees are compared to the true phylogeny (1a) using the selected error measure $\mathit{E}$, in our case the nLTT statistic [\cite{janzen2015approximate}]. Taking the differences between the true tree's nLTT and each posterior's nLTT results in an error distribution, here displayed as a histogram
    of error values (5a). Optionally, a twin phylogeny (1b) can be generated from the true phylogeny (1a), which then results in an error distribution (5b) with similar intermediate steps.
  }
  \label{fig:pipeline}
\end{figure}

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item for a given phylogeny an alignment is simulated under a known alignment model;
    \item for this alignment, according to the specified inference conditions, an inference model is picked (which may be different from the generative alignment model);
    \item the alignment is then used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error made, according to the error measure specified by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

>>>>

% Affiliations
\usepackage{authblk}

% RJCB: I think this title is better: 
% (1) tree prior importantance is the goal, the error is just the means
% (2) no need to specify BEAST2, it's also just a means
\title{pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics}
% \title{pirouette: Determining the error BEAST2 makes in inferring a phylogeny}

\author[1]{Rich\`el J.C. Bilderbeek}
\author[1]{Giovanni Laudanno}
\author[1]{Rampal S. Etienne}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}

@@

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will quantify the importance of a tree prior.

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & \textbf{Possible values} \\ 
    \hline
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model & Yule, BD, CBS, CCP, CEP \\
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates & strict, relaxed log-normal \\
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model & JC, HKY, TN93, GTR \\
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs & $m_{r} \in \mathbb{R}_{>0}$\\
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree & any combination of a, c, g, t \\
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common ancestor & see documentation \\
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup & see documentation \\
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model & Generative, Candidate \\
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is used & Always, Best candidate\\
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error & nLTT, $|\gamma|$ \\
    $\mathit{t_{m}}$ & twin tree prior & Sets the tree prior for the twin tree & Yule, BD \\
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. 
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian Skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and  Yano (\cite{hasegawa1985dating}), 
    TN93 = \cite{tamura1993estimation}, 
    GTR = Generalised time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a phylogeny (1a) simulated by generative prior $\mathit{P}$.
    The phylogeny is converted to an alignment (2a) using alignment model $\mathit{A}$. 
    The user supplies one or more experiments.
    For each experiment $\mathit{X}$ (a tuple of inference model $\mathit{I}$ and condition $\mathit{C}$),
    the inference model $\mathit{I}$ is used in the next step (3a),
    if the condition $\mathit{C}$ are satisfied.
    The condition $\mathit{C}$ can depend on the alignment (2a).
    The allowed inference models use the alignment (2a) 
    to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, a twin phylogeny (1b) can be generated from the original 
    phylogeny (1a), which then results in an error distribution (5b) 
    with similar intermediate steps.
  }
  \label{fig:pipeline}
\end{figure}

The pipeline is summarized by the following steps:
\begin{enumerate}
    \item for a given phylogeny an alignment is simulated under a known alignment model;
    \item for this alignment, according to the specified inference conditions, an inference model is picked (which may be different from the generative alignment model);
    \item the alignment is then used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error made, according to the error measure specified by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.
